---
title: "Week 7 Notes - Spatial ML and Spatial Autocorrelation"
date: "2025-10-20"
---

## Key Concepts Learned

### Types of Variables

**Continuous:** square feet, age of house, income levels, distance to downtown

**Categorical:** neighborhood, school district, building type, has garage (binary Y/N)

### Fixed Effects

-   Definition: Categorical variables that capture all unmeasured characteristics of a group

-   Powerful but they’re a *black box* 

-   Neighborhoods bundle many unmeasured factors (e.g. school districts, job access, amenities, "cool factor")

## Coding Techniques

-   **Make Data with X/Y Coordinates Spatial**

```{r, eval = FALSE}

```

-   **Add Dummy/Categorical Variables (e.g. Neighborhoods)**

The n-1 rule: One neighborhood is automatically chosen as the *reference category* (omitted)! R picks the first alphabetically unless you specify otherwise.

```{r, eval = FALSE}

```

-   **Interaction Effects**

```{r, eval = FALSE}
# 
```

-   **Polynomials – e.g. Square Age Variable**

```{r, eval= FALSE}
# 
```

-   **kNN**

```{r, eval = FALSE}
# 
```

## Questions & Challenges

-   

## Connections to Policy

-   

## Reflection
