---
title: "Week 3 Notes - Data Viz and Exploratory Analysis"
date: "2025-09-22"
---

## Key Concepts Learned

-   Data Viz basics with ggplot in RStudio
-   

## Coding Techniques

```{r ggplot}
# Formatting Charts
ggplot(demo_data) +
  aes(x = total_popE, y = median_incomeE) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Income vs Population in Pennsylvania Counties",
    subtitle = "2018-2022 ACS 5-Year Estimates",
    x = "Total Population",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau ACS"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::comma)
```

```{r}
# Finding Census Variable Codes
population_vars <- acs_vars_2022 %>%
  filter(str_detect(label, "Total.*population"))

age_vars <- acs_vars_2022 %>%
  filter(str_detect(label, "[Mm]edian.*age"))

housing_vars <- acs_vars_2022 %>%
  filter(str_detect(label, "[Mm]edian.*(rent|value)"))

```

## Questions & Challenges

-   With more reps, I hope to become more independent in my coding instead of having to refer to demo code

## Connections to Policy

-   When policymakers explore the use of algorithms, they need to ask themselves
    -   what proxy are they using?
    -   what are the possible blind spots?
    -   what are the resultant harms as well as potential guardrails?

## Reflection

-   It was really interesting to hear about the 2020 Census Innovation of Differential Privacy (adding mathematical noise as anomalies to protect respondents' privacy)
