---
title: "Week 5 Notes - Intro to Linear Regression"
date: "2025-10-06"
---

## Key Concepts Learned

1.  **The Statistical Learning Framework:** What are we actually doing?
2.  **Two goals:** Understanding relationships vs Making predictions
3.  **Building your first model** with PA census data
4.  **Model evaluation:** How do we know if it's any good?
5.  **Checking assumptions:** When can we trust the model?
6.  **Improving predictions:** Transformations, multiple variables

### **Statistical Significance**

**t-statistic:** How many standard errors away from 0?

-   Bigger \|t\| = more confidence the relationship is real

**p-value:** Probability of seeing our estimate if H₀ is true

-   Small p → reject H₀, conclude relationship exists

### Model Evaluation

-   **How well does it fit the data we used?** (in-sample fit using R²)

-   **How well would it predict new data?** (out-of-sample performance)

## Coding Techniques

-   **Linear Regression Train/Test Split**

```{r, eval = FALSE}
set.seed(123)
n <- nrow(pa_data)

# 70% training, 30% testing
train_indices <- sample(1:n, size = 0.7 * n)
train_data <- pa_data[train_indices, ]
test_data <- pa_data[-train_indices, ]

# Fit on training data only
model_train <- lm(median_incomeE ~ total_popE, data = train_data)

# Predict on test data
test_predictions <- predict(model_train, newdata = test_data)
```

-   **Evaluate Predictions**

```{r, eval = FALSE}
rmse_test <- sqrt(mean((test_data$median_incomeE - test_predictions)^2))
rmse_train <- summary(model_train)$sigma

cat("Training RMSE:", round(rmse_train, 0), "\n")
cat("Test RMSE:", round(rmse_test, 0), "\n")
```

-   **10-Fold Cross Validation**

```{r, eval= FALSE}
{r}
library(caret)

train_control <- trainControl(method = "cv", number = 10)

cv_model <- train(median_incomeE ~ total_popE,
                  data = pa_data,
                  method = "lm",
                  trControl = train_control)

cv_model$results
```

## Questions & Challenges

-   

## Connections to Policy

-   

## Reflection

-   
