---
title: "Week 3 Notes - Data Viz and Exploratory Analysis"
date: "2025-09-22"
---

## Key Concepts Learned

-   **Data Viz basics with ggplot in RStudio**
-   **Exploratory Data Analysis (EDA) Workflow**
    -   **Load and inspect** - dimensions, variable types, missing data
    -   **Assess reliability** - examine margins of error, calculate coefficients of variation
    -   **Visualize distributions** - histograms, boxplots for each variable
    -   **Explore relationships** - scatter plots, correlations
    -   **Identify patterns** - grouping, clustering, geographical patterns
    -   **Question anomalies** - investigate outliers and unusual patterns
    -   **Document limitations** - prepare honest communication about data quality
-   **Combining datasets through spatial joins**
    -   `left_join()` - Keep all rows from left dataset
    -   `right_join()` - Keep all rows from right dataset
    -   `inner_join()` - Keep only rows that match in both
    -   `full_join()` - Keep all rows from both datasets

## Coding Techniques

**Formatting Charts: Grammar of Graphics**

Data → Aesthetics → Geometries → Visual

```{r ggplot, eval = FALSE }
# Formatting Charts
ggplot(demo_data) +
  aes(x = total_popE, y = median_incomeE) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Income vs Population in Pennsylvania Counties",
    subtitle = "2018-2022 ACS 5-Year Estimates",
    x = "Total Population",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau ACS"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::comma)
```

**Finding Census Variable Codes**

```{r census, eval = FALSE}
# Finding Census Variable Codes
population_vars <- acs_vars_2022 %>%
  filter(str_detect(label, "Total.*population"))

age_vars <- acs_vars_2022 %>%
  filter(str_detect(label, "[Mm]edian.*age"))

housing_vars <- acs_vars_2022 %>%
  filter(str_detect(label, "[Mm]edian.*(rent|value)"))

```

## Questions & Challenges

-   With more reps, I hope to become more independent in my coding instead of having to refer to demo code

## Connections to Policy

-   Summary statistics can hide critical patterns
-   Outliers may represent important communities
-   Relationships aren’t always linear
-   Visual inspection reveals data quality issues

## Reflection

-   I really liked the Anscombe's Quartet example – it is a really powerful and intuitive example to show the power of data visualization
-   Astounding Statistic that only **27% of planners** warn users about unreliable ACS data – I admit to being guilty of this oversight too
